### 2015-04-13 motivation.tex
rework gaps.

Recently, a number of package managers have been developed to manage 
multi-configuration builds.
%
Nix~\cite{dolstra+:icfp08,dolstra+:lisa04}
is a package manager and an OS distribution that supports the installation of
arbitrarily many software configurations.  As at most HPC sites, each package
in Nix installed in a unique prefix, but in Nix, there is no human-readable
naming convention.  Instead, Nix determines the prefix by hashing the package
file and all of its dependencies.
%
Hashdist~\cite{hashdist} is a meta-build system and package manager for HPC.
Like Nix, it uses cryptographic versioning, and it archives installations
in a git repository so that they can be recalled on demand. 
%
Both Nix and hashdist use {\tt RPATHs} in their packages to ensure that
libraries are found correctly.

The EasyBuild~\cite{hoste+:pyhpc12} tool is in production use at
the University of Ghent.  It allows multiple versions to be installed at once,
but it requires package authors to manage combinatoric package
versioning. Rather than setting {\tt RPATHs}, it auto-generates module files
to manage each package's environment, and it is closely coupled with 
Lmod~\cite{geimer+:hust14}.  EasyBuild groups
the compiler, MPI, FFT, and BLAS libraries together in a
{\it toolchain}, providing limited package composability. 
Smithy~\cite{digirolamo:smithy} is a similar tool in use at ORNL. It
can generate module files, but it does not provide any 
automated dependency management; it only checks whether a package's
prerequisites have already been installed by the user.

We observed a number of shortcomings in these existing systems. While the
cryptographic versioning of Nix and Hashdist is appealing, it does not allow
users to easily navigate the installed software or to find packages consistent
with their own setup.  In fact, the only way to access software after it is built
is to recall it by its hash. EasyBuild and Smithy address some of these problems
by providing modules for the installed packages, but neither has a good system
for managing the full space of combinatoric names.  EasyBuild attempts to do this by
lumping versions of common packages together in a tooclhain, but the toolchains
themselves have cryptic names like {\tt goolf} 
(\textbf{g}cc, \textbf{o}penmpi, \textbf{o}penblas, \textbf{S}caLAPACK,
\textbf{F}FTW), and their versions are meaningless. {\tt RPATH} handling, where
provided, is left up to the package authors, and is not consistently enforced.

The main limitation of existing tools is the lack of build {\it composability}.
HPC configuration space is combinatoric, and realizing an arbitrary
combination of compilers, MPI version, build options, and dependency versions
requires tediously modifying many packages implementation.  
No existing system parameterizes 
package files to deal with this issue.  Indeed, the number of package files
required for existing systems scales with the number of version {\it combinations},
not the number of packages.  This quickly becomes unmanageable.  Indeed, EasyBuild
maintains over 3,300 files for only 600 packages.  Other systems have similar
issues but limit the number of supported configurations.

>>>>

Recently, a number of package managers have been developed to manage 
multi-configuration builds.
%
Nix~\cite{dolstra+:icfp08,dolstra+:lisa04}
is a package manager and an OS distribution that supports the installation of
arbitrarily many software configurations.  As at most HPC sites, each package
in Nix installed in a unique prefix, but in Nix, there is no human-readable
naming convention.  Instead, Nix determines the prefix by hashing the package
file and along with its dependencies.

Hashdist~\cite{hashdist} is a meta-build system and package manager for HPC.
Like Nix, it uses cryptographic versioning, and it archives installations
in a git repository so that they can be recalled on demand. 
%
Both Nix and hashdist use {\tt RPATHs} in their packages to ensure that
libraries are found correctly.

The EasyBuild~\cite{hoste+:pyhpc12} tool is in production use at
the University of Ghent.  It allows multiple versions to be installed
at once.  Rather than setting {\tt RPATHs}, it auto-generates module files
to manage each package's environment, and it is closely coupled with 
Lmod~\cite{geimer+:hust14}.  EasyBuild groups the compiler, MPI, FFT, and
BLAS libraries together in a {\it toolchain}, which can be combined with
any package file. This provides some degree of composability and
separates compiler flags and MPI concerns from client packages.

Smithy~\cite{digirolamo:smithy} is a similar tool in use at ORNL. It
can generate module files, but it does not provide any 
automated dependency management; it only checks whether a package's
prerequisites have already been installed by the user.

\paragraph{Gaps in the state of the art}
The cryptographic versioning of Nix and Hashdist is very flexible.  It versions
the package {\it and} its dependency configuration, and can represent any
configuration. However, there is no way for users to navigate or query
software. The systems are in a sense ``write-only'':
Nix does not even offer a way to ask for a package's dependencies.
%
EasyBuild and Smithy generate environment modules, which allow querying
through their own interface. Naming in existing module systems, however, cannot
handle combinatoric versions. The Lmod authors call this the
``matrix problem``~\cite{mclay:lmod-tutorial}.

%EasyBuild's attempts to version groups dependencies by adding versions to
%toolchains, but the naming is difficult to understand.
%. Compiler, MPI, and 
%some library versions are lumped together, but the results is cryptic, e.g.
%{\tt goolf} stands for ``gcc, openmpi, openblas, ScaLAPACK, FFTW'', and
%its version is meaningless.

Existing tools do not enforce {\tt RPATHs}, leaving them to package authors.
This may lead to erroneous runs.

The main limitation of existing tools is the lack of build {\it composability}.
The full set of package versions is combinatoric, and realizing an arbitrary
combination of compiler, MPI version, build options, and dependency versions
requires tediously modifying many package files.
%
Indeed, the number of package files required for existing systems scales with
the number of version {\it combinations}, not the number of packages.
This quickly becomes unmanageable.  As an example, the very flexible EasyBuild
system has over 3,300 files for several permutations of only 600 packages. 
Other systems have similar issues. A mechanism is needed to {\it parameterize}
packages so that new builds can be {\it composed} on the fly.
